---
title: "medicalcoder:"
subtitle: "A Unified and Longitudinally Aware Framework for ICD-Based Comorbidity Assessment in R"
author:
  - name: "Peter E. DeWitt, Ph.D."
    affiliation:
      - "University of Colorado Anschutz, <br>Department of Biomedical Informatics"
date: "May 8, 2026"
data-format: "MMMM D, YYYY"
format:
  revealjs:
    theme: 
      - default
      - slides.scss
    fontsize: 2em
    html-math-method: mathjax
    slide-number: c
    show-slide-number: all
    highlight-style: rstudio_default-light.theme
embed-resources: true   # true = a single file, false = multiple files
knitr:
  opts_chunk:      ########## set global options ############
    collapse: true # keep code from blocks together (if shown)
    echo: false    # don't show code
    message: true  # show messages
    warning: true  # show warnings
    error: true    # show error messages
    R.options:   
      digits: 3    # round to three digits
editor_options:
  chunk_output_type: console
editor: source
highlight-style: rstudio_default-light.theme
---

```{r}
#| label: setup
#| inlclude: false
# load data
options("datatable.print.topn" = 3)

mimiciv <- arrow::read_feather("../mimiciv.feather")
data.table::setDT(mimiciv)
```
 
# Introduction

## Comorbidity algorithms based on International Classification of Diseases (ICD)

* Used for:
  * Risk adjustment, and
  * Cohort characterization.

* Existing implementations
  * fragment across packages,
  * inconsistently handle mixed ICD-9 and ICD-10 data
  * encounter-level logic only

## Longitudinal Awareness

* Existing implementations are encounter-level: 

  __This ignores the longitudinal nature of medical records!__

* Encounter-level flagging is sufficient if:

  * _all_ ICD codes from prior encounters are present in the current encounter
    record, and
  * primary diagnosis flags are correct
  * present-on-admission flags are correct

* Simple carry-forward of encounter-level flagging within a patient record is
  insufficient
  * Disease severity can be missed
  * Some conditions will be missed entirely

## We present the R package __medicalcoder__

:::: {.columns}

::: {.column width="65%"}

* Comorbidity Flagging Options:
  * Encounter-level
  * __Longitudinally aware__

* Accommodates:
  * Present-on-admission flags
  * Primary diagnosis flags
  * Mixed ICD versions
  * Mixed full and compact ICD code formats

* multiple variants of 
  * Charlson, 
  * Elixhauser, and 
  * Pediatric Complex Chronic Conditions (PCCC)

:::

::: {.column width="35%"}

```{r}
#| label: hexlogo
#| echo: false
knitr::include_graphics(
  system.file("help", "figures", "hex.svg", package = "medicalcoder")
)
```

:::

::::

# Examples

## Example Data: MIMIC-IV v3.1

* De-identified health data from patients admitted to intensive care
  units at Beth Israel Deaconess Medical Center.

* ICD-9 and ICD-10 diagnostic and procedure codes available for comorbidity
  research.

* MIMIC-IV is widely used for method development, validation, and reproducible
  research in clinical informatics and outcomes research.

* https://physionet.org/content/mimiciv/3.1/


## Example Data: MIMIC-IV v3.1

```{r}
#| label: mimiciv-data
#| echo: true
mimiciv
```

::: {style="font-size: 0.7em;"}

| Column         | Description                                        | Source                | 
| :--            | :------                                            | :--                   | 
| `subject_id`   | subject identifier                                 | MIMIC-IV              | 
| `encounter`    | Encounter Secquence within `subject_id`            | constructed           | 
| `age`          | Age, in years, at admission                        | MIMIC-IV; constructed | 
| `icd_code`     | ICD code                                           | MIMIC-IV              | 
| `icd_version ` | ICD version integer value 9 or 10                  | MIMIC-IV              | 
| `dx`           | Indicator for diagnostic (1) or procedure (0) code | MIMIC-IV; constructed | 
| `poa`          | Present-on-admission flag                          | constructed           | 
| `pdx`          | Primary diagnosis flag                             | constructed           | 

:::
